---
title: "Psych 434: Conducting Research Across Cultures"
author:
    name: Joseph Bulbulia
    orcid: 0000-0002-5861-2056
    affiliation: Victoria University of Wellington, New Zealand
    email: joseph.bulbulia@vuw.ac.nz
    corresponding: yes
    echo: false
    warnings: false
    error: false
    highlight: kate
    code-fold: true
    code-overflow: scroll
    code-line-numbers: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
# Necessary for using dvisvgm on macOS
# See https://www.andrewheiss.com/blog/2021/08/27/tikz-knitr-html-svg-fun/
# Sys.setenv(LIBGS = "/usr/local/share/ghostscript/9.53.3/lib/libgs.dylib.9.53")
# font_opts <- list(dvisvgm.opts = "--font-format=woff")
```

```{r}
#| echo: false
library("tinytex")
```

# Welcome to the Psych 434 lecture website



::: {.callout-tip}
### Accessing Lectures and Readings

- **Seminar Time/Location**: Mondays, 11:00-12:50pm **Easterfield Building Room 406**

- **Course Outline**: find a detailed schedule of topics, readings, and assignments in the `Course Outline` tab.

- **Readings**: links to readings are directly within the `Course Outline` tab, essential for lecture preparation.

- **Lecture Materials**: access slides, video recordings, and more under the `Content` tab, organised by week for ease of use.

- **Tests**: in the same room as the seminar
:::



## Class times and locations

**Seminar** Day/Time: Mondays, 11:00-12:50pm

**Seminar Location** Easterfield Building EA406

**Test/Quiz Location** Easterfield Building EA406

## Contents Tab

The `Contents` tab offers direct access to weekly seminar and lab materials, including lecture outlines and lab resources.

- Access it from the top right of the course platform by selecting the appropriate week.

- Lab materials are available one week before the lecture; seminar review materials post-seminar.


```{r .column-body-outset}
#| echo: false
library(leaflet)
leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=174.7681, lat=-41.2889, popup="Easterfield")
```

## Names and contact details

**Course Coordinator** Prof Joseph Bulbulia [joseph.bulbulia\@vuw.ac.nz](mailto:joseph.bulbulia@vuw.ac.nz){.email}

**Course Coordinator's Office** EA324

**Teaching Fellow** Dr.Inkuk Kim [inkuk.kim\@vuw.ac.nz](mailto:inkuk.kim@vuw.ac.nz){.email}

**Laboratory Assistant** Dr Torven Schalk [torven.schalk\@vuw.ac.nz](mailto:torven.schalk@vuw.ac.nz){.email}




## Assignments and due dates

```{r }
#| echo: false
#| code-fold: true
#| warnings: false
#| message: false

library(tidyverse)
library(knitr)
library(kableExtra)

# create data frame
my_data <- tibble(
  Assessment = c(
    "Class participation",
    "Test 1",
    "Test 2",
    "Research Report"
  ),
  CLOs = c("1,2,3", "2", "2", "1,2,3"),
  Percent = c("10", "25", "25", "40"),
  Due = c(
    "Weekly",
     "25 March (w5)",
     "6 May (w9)",
     "3 June (w12)"
  )
  
)

# create table 
my_data %>%
  kbl(format = "html") |> 
  kable_styling("hover")

```


## Course Description

The official description:

*This course will focus on theoretical and practical challenges for conducting research involving individuals from more than one cultural background or ethnicity. Topics are likely to include defining and measuring culture; developing culture-sensitive studies, choice of language and translation; communication styles and bias; questionnaire and interview design; qualitative and quantitative data analysis for cultural and cross-cultural research; minorities, power and ethics in cross-cultural research; and ethno-methodologies and indigenous research methodologies. Appropriate background for this course: PSYC 338.*

## Course Learning Objectives

**Preamble:** in this advanced course, students will develop foundational skills in cross-cultural psychological research with a strong emphasis on causal inference, a new and critical methodological approach. 

1.  **Programming in R** students will learn the basics of programming in the statistical language R, gaining essential computational tools for psychological research. The skills you acquire will lay the foundation for applying data analysis techniques in a causal inference framework and beyond.

2.  **Understanding Causal Inference.** students will develop a robust understanding of causal inference concepts and approaches, with particular emphasis on how they mitigate common pitfalls in cross-cultural research. We will focus on designing studies, analysing data, and drawing strong conclusions about cause-and-effect relationships across cultures.

5.  **Understanding Measurement in Comparative Settings.** students will learn techniques for constructing and validating psychometrically sound measures across diverse cultures. We will examine how to ensure measurements are reliable, cross-culturally valid, and aligned with theoretical constructs while focusing strongly on causal reasoning. 

## Assignments and due dates

```{r }
#| echo: false
#| code-fold: true
#| warnings: false
#| message: false

library(tidyverse)
library(knitr)
library(kableExtra)

# create data frame
my_data <- tibble(
  Assessment = c(
    "Class participation",
    "Test 1",
    "Test 2",
    "Research Report"
  ),
  CLOs = c("1,2,3", "2", "2", "1,2,3"),
  Percent = c("10", "25", "25", "40"),
  Due = c(
    "Weekly",
     "25 March (w5)",
     "6 May (w9)",
     "3 June (w12)"
  )
  
)

# create table 
my_data %>%
  kbl(format = "html") |> 
  kable_styling("hover")

```

### Assessment 1: Class Participation

- **Attendance and Participation**:
    - (a) Class attendance and active participation.
    - (b) Lab attendance and active participation.

### Assessment 2: In-class Tests

::: {.callout-important}
## Test Guidelines

- Test duration is under an hour. The allocated time is nearly two hours.
- Each lecture starts and ends with key concept definitions and reviews for the test.
- R or RStudio knowledge isn't part of the test. R support aims to enhance research report skills.
- Tests are conducted in the lecture room.
- Required: pen/pencil. Computers are not allowed.
:::
- **First Test (50 minutes, total time allowed: 1 hour 50 minutes)**:
    - Focuses on revising core statistical and methodological concepts.
    - Aims to refresh basic statistical knowledge foundational for later course material.

- **Second Test (50 minutes, total time allowed: 1 hour 50 minutes)**:
    - Builds upon the first test's concepts, emphasising the application of basic conceptual, statistical, and theoretical knowledge.

### Assessment 3: Research Report

::: {.callout-important}
## Research Report Instructions
- We will supply the data.
- Lab sessions are designed to support you in this assignment.
- We assume no statistical background. 
:::

- **Title**: "Causal Inference in Cultural Psychology: Examining Exposure Effects on Dimensions of Well-being Modified by Cultural or Sociodemographic Categories".

- **Objective**:
    - To quantify the causal effect of a specific exposure on well-being dimensions, modified by sociodemographic categories (`born_nz`, `eth_cat`, `big_doms`, `gen_cohort`) using the NZAVS longitudinal synthetic dataset.

- **Instructions**:

  1. **Theoretical Interest and Research Question**:
        - Describe the significance of your chosen exposure and its potential impact on the selected outcomes, modified by the cultural or sociodemographic category.
        - State the research question clearly.

  2. **Directed Acyclic Graph (DAG)**:
        - Construct a DAG illustrating the relationships between exposure, outcomes, sociodemographic category, and potential bias sources. Ensure clarity in labelling.

  3. **Confounding Control Strategy**:
        - Outline your strategy for confounding control, justifying the chosen confounders.

  4. **Measurement Biases**:
        - Address and analyse measurement biases as relevant.

    5. **Assumptions and Statistical Models**:
        - Discuss the assumptions of your causal inference approach and your statistical model, including their limitations.

- **Requirements**:
    - Introduction: 1,500 words **limit.**
    - Conclusion: 1,500 words  **limit.**
    - Method and Results sections should be concise; no specific word limit.
    - Use any useful sources, citing appropriately to avoid academic misconduct.
    - Follow APA style for citations and references.
    - Include tables/figures as needed.
    - Submit as a single PDF, including R code in an appendix.

- **Evaluation Criteria**:
    - Clarity of theoretical framework, research question, and design.
    - Validity of confounding control strategy.
    - Discussion on assumptions and statistical models.
    - Organisation and presentation quality.

### Extensions and Penalties

- **Extensions**:
    - Negotiate a new due date by writing (email) before **March 10th, 2024**, if necessary.

- **Penalties**:
    - Late submissions incur a one-grade-per-day penalty.
    - Over-length assignments will be penalised.

- **Unforeseeable Events**:
    - Extensions post-March 10 require evidence (e.g., medical certificate).

### Materials and Equipment

- Bring a laptop with R and RStudio installed for data analysis sessions. Contact the instructor if you lack computer access.
- For in-class tests, bring a writing utensil. Electronic devices are not permitted.



